---
title: "Moody's Analytics Coding Test"
author: "Said Maanan"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Introduction

In this document I present my solution to the Moody's Analytics Coding Test for Candidates. 

## Problem 1

The first step to answering the questions in the first problem is to load the data frame `data.csv`:

```{r}
library(readr)
data <- read_csv("~/Downloads/Coding_Test/data.csv")
```

The `readr` package tells us that the data frame consists of 5 variables and 13646 observations. Now we can start answering our questions.

### Question 1

Generate the variable `age` for each loan:

```{r}
library(lubridate)
data$age = - round(interval(today(), ymd(data$time, truncated = 2)) / years(1))
```

What is the maximum age of the default rate observed in the sample?

```{r}
max(data$age)
```

The maximum age of non-missing default rate observed in the sample is 17 years.

### Question 2

Create a table reporting the summary statistics for `default_rate`:

```{r}
summary(data$default_rate)
```

Percentage of missing cells:

```{r}
sum(is.na(data$default_rate))/length(data$default_rate)*100
```

### Question 3

Generate the variable `lifecycle`:

```{r}
library(tidyverse)
```

```{r}
data <- data %>% 
  group_by(age) %>% 
  mutate(lifecycle = mean(default_rate, na.rm = TRUE))
```

Plot `lifecycle` over age using line chart:

```{r}

plot(unique(data[,c("age","lifecycle")]), type = "b", pch = 19, col = "red",
     main = "Age vs Lifecycle",
     xlab = "Age",
     ylab = "Lifecycle")
```

### Question 4

Pick two loans with the default rate observed for at least five periods.

```{r}
book_1 = data[data$account_id=='400293',]
book_2 = data[data$account_id=='401749',]
```

For each loan, plot the lifecycle of the default rate.

```{r}
par(mfrow=c(1,2))

plot(x = book_1$age, y = book_1$default_rate, type = "b", pch = 19, col = "red",
     main = "Loan 1",
     xlab = "Age",
     ylab = "Default rate")

plot(x = book_2$age, y = book_2$default_rate, type = "b", pch = 19, col = "red",
     main = "Loan 2",
     xlab = "Age",
     ylab = "Default rate")
```

Do the two loans follow the portfolio's lifecycle?

```{r}

par(mfrow=c(1,2))

plot(x = book_1$age, y = book_1$default_rate, type = "b", pch = 19, col = "red", ylim = c(0, 0.035),
     main = "Loan 1",
     xlab = "Age",
     ylab = "Default rate")

points(unique(data[,c("age","lifecycle")]), type = "p", pch = 19, col = "black")

lines(unique(data[,c("age","lifecycle")]), lty = 2, col="black")

plot(x = book_2$age, y = book_2$default_rate, type = "b", pch = 19, col = "red", ylim = c(0, 0.08),
     main = "Loan 2",
     xlab = "Age",
     ylab = "Default rate")

points(unique(data[,c("age","lifecycle")]), type = "p", pch = 19, col = "black")

lines(unique(data[,c("age","lifecycle")]), lty = 2, col="black")
```

The first account `400293` seems to align closely with the portfolio's lifecycle, but the second account `401749` doesn't.

### Question 5

Estimate a fixed-effect model for the `default_rate` as target variable and `age` and `r12_gdp` as fixed effects:

```{r}
library(fixest)
mod_bl <- feols(default_rate ~ account_id | age^r12_gdp_bl, data = data)
summary(mod_bl)

mod_st <- feols(default_rate ~ account_id | age^r12_gdp_bl, data = data)
summary(mod_st)
```

Obtain predictions

```{r}
fitted_bl = mod_bl$fitted.values
fitted_st = mod_st$fitted.values
```

How can you ensure that the predictions are bounded between 0 and 1?

In order to ensure that the predictions are bounded between 0 and 1, it is more appropriate to use a Fixed-Effect Generalized Linear Model instead of an Ordinary Least Squares.

```{r}
mod_bound = feglm(default_rate ~ account_id | age + r12_gdp_bl, data = data, family = "quasibinomial")
summary(mod_bound$fitted.values)

mod_bound_st = feglm(default_rate ~ account_id | age + r12_gdp_st, data = data, family = "quasibinomial")
```


Plot the actual vs the predicted values of the default rate for the two loans selected previously.

```{r}
pred_1 = predict(mod_bound, book_1)
pred_2 = predict(mod_bound, book_2)

df_1 = data.frame(Age = book_1$age, Actual = book_1$default_rate, Predicted = pred_1)
df_2 = data.frame(Age = book_2$age, Actual = book_2$default_rate, Predicted = pred_2)

par(mfrow=c(1,2))

plot(x = df_1[-1,]$Actual, y = df_1[-1,]$Predicted, type = "p", pch = 19, col = "black",
     main = "Loan 1",
     xlab = "Actual",
     ylab = "Predicted")

plot(x = df_2[-1,]$Actual, y = df_2[-1,]$Predicted, type = "p", pch = 19, col = "black",
     main = "Loan 2",
     xlab = "Actual",
     ylab = "Predicted")

```

Aggregate actual rates and predicted rates by age.

```{r}
par(mfrow=c(1,2))

plot(x = df_1$Age, y = df_1$Predicted, type = "p", pch = 19, col = "red",
     main = "Loan 1",
     xlab = "Time",
     ylab = "Default rate")

points(x = df_1$Age, df_1$Actual, type = "p", pch = 19, col = "black")

legend("topright", c("Actual","Predicted"), col=c("black","red"), pch=c(19,19))

plot(x = df_2$Age, y = df_2$Predicted, type = "p", pch = 19, col = "red",
     main = "Loan 2",
     xlab = "Time",
     ylab = "Default rate")

points(x = df_2$Age, df_2$Actual, type = "p", pch = 19, col = "black")

legend("topright", c("Actual","Predicted"), col=c("black","red"), pch=c(19,19))
```

## Problem 2

### Question 1

```{r}
library(readr)
mortgages_small <- read_csv("~/Downloads/Coding_Test/mortgages_small.csv")
```

How many observations in the data set?

```{r}
nrow(mortgages_small)
```

How many loans in the data set?

```{r}
length(levels(as.factor(mortgages_small$loan_id)))
```

What is the time coverage in terms of months?

```{r}
library(lubridate)

min(ym(mortgages_small$obs_month))

max(ym(mortgages_small$obs_month))
```

### Question 2

Create a table reporting the number of loans and percentage of loans per month.

```{r}
mortgages_small$obs_m = ym(mortgages_small$obs_month)
mortgages_small$month = month(mortgages_small$obs_m)

monthly_loans = mortgages_small %>%
  group_by(month) %>%
  dplyr::summarise(n = n())

monthly_loans$prc = monthly_loans$n/sum(monthly_loans$n)*100

monthly_loans
```

Create a bar plot showing the number of loans at each observation date.

```{r}

barplot(monthly_loans$n,
        main = "Loans vs Months",
        xlab = "Months",
        ylab = "Loans",
        names.arg = 1:12)
```

### Question 3

Create a table reporting the number and share of observations per origination year.

```{r}
mortgages_small$ori_y = ym(mortgages_small$orig_date)
mortgages_small$year = year(mortgages_small$ori_y)

yearly_loans = mortgages_small %>%
  group_by(year) %>%
  dplyr::summarise(n = n())

yearly_loans$prc = yearly_loans$n/sum(yearly_loans$n)*100

yearly_loans
```

Create a bar plot showing the number of loans at each origination date.

```{r}
barplot(yearly_loans$n,
        main = "Loans vs Years",
        xlab = "Years",
        ylab = "Loans",
        names.arg = yearly_loans$year)
```

### Question 4

Create a new variable that is the first observation date of each account.

```{r}
library(dplyr)

mortgages_small =  mortgages_small %>%
  group_by(loan_id) %>%
  mutate(first_obs = min(obs_m)) %>%
  ungroup()
```

Create and summarize the observation lag.

```{r}
mortgages_small$obs_lag = - interval(ymd(mortgages_small$obs_m), ymd(mortgages_small$first_obs)) / days(1)
summary(mortgages_small$obs_lag)
```

### Question 5

Merge `macro.csv` and `mortgages.csv`

```{r}
macros <- read_csv("~/Downloads/Coding_Test/macros.csv")
macros$date = ym(macros$date)
merged = merge(x = mortgages_small, y = macros, by.x = "obs_m", by.y = "date", all = TRUE)
```

### Question 6

Create an indicator variable `default_flag`.

```{r}
merged = mutate(merged, default_flag = ifelse(arrears >= 3, 1, 0))
```

### Question 7

Create a single table reporting the summary statistics.

```{r}
merged %>% 
    select(default_flag, current_balance, current_ltv, credit_score, original_ltv, hpi_o, hpi_t) %>% 
    map_df(.f = ~ broom::tidy(summary(.x)), .id = "variable")
```

### Question 8

Tabulate the variables `arrears`, `mortgage_type` and `occupancy`.

```{r}
library(janitor)
df = data_frame(merged$arrears, merged$mortgage_type, merged$occupancy)
apply(df, 2, tabyl)
```

### Question 9

Remove missing observations from the data set.

```{r}
merged$d_hpi = merged$hpi_t - merged$hpi_o
merged = na.omit(merged)
```

### Question 10

Partition the data set into partition and test samples

```{r}
library(caret)
set.seed(1219)

merged$default_flag = as.factor(merged$default_flag)
ind   = createDataPartition(merged$mortgage_type, p=0.8,list = F)
train = merged[ind,]
test  = merged[-ind,]
```

### Question 11

Estimate a logit model.

```{r}
log_mod = glm(default_flag ~ credit_score + occupancy + current_ltv + mortgage_type + gdp_t + d_hpi, data = train, family = binomial)
```

### Question 12

Interpret the coefficients of the model.

```{r}
summary(log_mod)
```

### Question 13

Generate predicted default rates from the model.

```{r}
train$pred_default = log_mod$fitted.values
```

### Question 14

In-sample model validation.

Plot average predicted vs average actual over time.

```{r}
train$default_flag = as.numeric(as.character(train$default_flag))
agg_1 = train %>% group_by(obs_m) %>% summarise_at(vars(default_flag, pred_default), mean)

plot(agg_1$obs_m, agg_1$default_flag,
     main = "Actual Default vs Predicted Default",
     xlab = "Time",
     ylab = "Default rate",
     type = "o", col  = "blue", pch=20,
     )
lines(agg_1$obs_m, agg_1$pred_default, 
      lty = 2, col="red")

points(agg_1$obs_m, agg_1$pred_default,
       col="red", pch = 18)

legend("topright", c("Actual","Predicted"), lty = c(1,2), col=c("blue","red"), pch=c(20,18))
```

Create 3 categories of borrower credit score.

```{r}
train = mutate(train, bcs = ifelse(credit_score<= 650, "low-score", ifelse(credit_score > 650 & credit_score <= 800, "mid-score", "high-score")))

agg_2 = train %>% group_by(bcs) %>% summarise_at(vars(default_flag, pred_default), mean)

barplot(t(as.matrix(agg_2[,-1])),
        main = "Actual Default vs Predicted Default",
        xlab = "Score Category",
        ylab = "Default Rate",
        names.arg = c("High Score", "Low Score", "Mid Score"),
        col = gray.colors(2),
        beside = TRUE)
legend("topleft",
       c("Default", "Pred. Default"),
       fill = gray.colors(2))
```

### Question 15

Generate out-of-sample predictions.

```{r}
test$oos_pred = predict(log_mod, test, type = "response")

test$default_flag = as.numeric(as.character(test$default_flag))
```

Plot average predicted vs average actual over time.

```{r}

agg_3 = test %>% group_by(obs_m) %>% summarise_at(vars(default_flag, oos_pred), mean)

plot(agg_3$obs_m, agg_3$default_flag,
     main = "Actual Default vs Predicted Default",
     xlab = "Time",
     ylab = "Default rate",
     type = "o", col  = "blue", pch=20,
     )
lines(agg_3$obs_m, agg_3$oos_pred, 
      lty = 2, col="red")

points(agg_3$obs_m, agg_3$oos_pred,
       col="red", pch = 18)

legend("topright", c("Actual","Predicted"), lty = c(1,2), col=c("blue","red"), pch=c(20,18))
```

### Question 16

Create `outcome` variable.

```{r}
test = mutate(test, outcome = ifelse(oos_pred > 0.1, 1, 0))

# test$outcome = as.factor(test$outcome)
```


### Question 17

Create confusion matrix.

```{r}
library(InformationValue)
optimal = optimalCutoff(test$default_flag, test$outcome)[1]
confusionMatrix(test$default_flag, test$outcome)
```

### Question 18

Create a receiver operating curve.

```{r}
library(PRROC)
PRROC_obj <- roc.curve(scores.class0 = test$oos_pred, weights.class0=test$default_flag, curve=TRUE)
plot(PRROC_obj)
```


